---
title: "UKRR data wrangling"
author: "Sally Thompson"
format: 
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual

execute:
  echo: false
  output: true
  warning: false
---

```{r setup}

library(here)

source("functions helpers.R")

options(scipen = 999)
```

## Intro

Code to extract quarterly counts and new cases per quarter to KRT services, from UKRR data extract.

Steps

- Read in UKRR incidence and prevalence extracts
- Create list of start and end dates for quarters, from Jan 2022
- Prevalence - filter those who died before Jan 2022
- For both files:
  - For each quarter, count who started before quarter-start and died after quarter-end - by region

  - Count those who started after quarter-start but before quarter-end - by region and referral

```{r read-data}

raw_timeline <- read_csv(here("data", "timeline_final24", "timeline_final24.csv")) |> 
  group_by(study_id) |> 
  slice_tail(n = 1) |>
  ungroup() |>
  filter(ttmod >= 92) |> 
  mutate(td_date = dmy(tdstart)) |>  
  select(study_id, td_date, ttmod)

raw_incident <- read_csv(here("data", "timeline_final24", "incident_demog24.csv"))|> 
  mutate(date_death = parse_date(dodeath, "%d%b%Y"),
         date_start = parse_date(datestart, "%d%b%Y")) |> 
  dplyr::select(study_id, flag, date_death, date_start, referral, region) |> 
  left_join(raw_timeline) |> 
  mutate(date_death = case_when(ttmod >= 92 ~ td_date),
         referral = replace_na(referral, "Early"))


raw_prev <- read_csv(here("data", "timeline_final24", "prev09_demog24.csv")) |>
  mutate(date_death = parse_date(dodeath, "%d%b%Y"),
         date_start = parse_date(datestart, "%d%b%Y")) |> 
  dplyr::select(study_id, flag, date_death, date_start, region) |> 
  left_join(raw_timeline) |> 
  mutate(date_death = case_when(ttmod >= 92 ~ td_date))


```

## UKRR vs NHS regions

Check whether the regions/region names match.

```{r}

ukrr_regions <- raw_incident |> 
  distinct(region) |> 
  arrange(region)
```

UKRR has separate regions for North East and "Yorkshire and Humber", whereas NHS combines these to form "North East and Yorkshire".

```{r}

raw_incident <- raw_incident |> 
  mutate(region = case_when(region %in% c("North East", "Yorkshire & Humber") ~ "North East and Yorkshire",
                             TRUE ~ region)) |> 
  filter(!region %in% c("N Ireland", "Scotland", "Wales", "zNI", "zSco", "zWls"))
  
raw_prev <- raw_prev |> 
  mutate(region = case_when(region %in% c("North East", "Yorkshire & Humber") ~ "North East and Yorkshire",
                             TRUE ~ region)) |> 
  filter(!region %in% c("N Ireland", "Scotland", "Wales", "zNI", "zSco", "zWls"))
```

## Flow counts (incidence)

Take incidence data, floor the start date to start of quarter then count.

Incidence up to 2024-Q4 (i.e. new cases up to end of 2024).

```{r incidence}

incid_quart <- raw_incident |> 
  mutate(new_quart = floor_date(date_start, unit = "quarter")) |> 
 # filter(new_quart >= "2021-01-01") |> 
  group_by(region) |> 
  count(new_quart) |> 
  ungroup() 
  

incid_eng <- incid_quart |> 
  group_by(new_quart) |> 
  summarise(n = sum(n)) |> 
  ungroup() |> 
  mutate(region = "England")

incid_quart <- bind_rows(incid_quart, incid_eng)

incid_qrt_total <- incid_quart |> 
  group_by(region, new_quart) |> 
  summarise(n = sum(n))


mort_quart <- raw_incident |> 
  mutate(mort_quarter = floor_date(date_death, unit = "quarter")) |> 
  filter(!(is.na(date_death))) |> 
  group_by(region) |> 
  count(mort_quarter) |> 
ungroup()

mort_eng <- mort_quart |> 
  group_by(mort_quarter) |> 
  summarise(n = sum(n)) |> 
  ungroup() |> 
  mutate(region = "England")

mort_quart <- bind_rows(mort_quart, mort_eng)



## alternative for stella flows - report by year

incid_yr <- raw_incident |> 
  mutate(new_year = floor_date(date_start, unit = "year")) |> 
 # filter(new_quart >= "2021-01-01") |> 
  group_by(region) |> 
  count(new_year) |> 
  ungroup()

incid_yr_eng <- incid_yr |> 
  group_by(new_year) |> 
  summarise(n = sum(n)) |> 
  ungroup() |> 
  mutate(region = "England")

incid_yr <- bind_rows(incid_yr, incid_yr_eng)



mort_yr <- raw_incident |> 
  mutate(mort_year = floor_date(date_death, unit = "year")) |> 
  filter(!(is.na(date_death))) |> 
  group_by(region) |> 
  count(mort_year) |> 
ungroup()

mort_yr_eng <- mort_yr |> 
  group_by(mort_year) |> 
  summarise(n = sum(n)) |> 
  ungroup() |> 
  mutate(region = "England")

mort_yr <- bind_rows(mort_yr, mort_yr_eng)
```

## Prevalence

Follow UKRR and CVD Prevent reporting of count on a census day. Take first day of each quarter as count day, apply to start value for that time period.

Prevalence to end of 2024-Q4 (up to end of year).

Need to add end-date, which is either date of death or date after the end of the data extract (which runs to 31 December 2024).

Calibrating in different SD model, starting at 2018 (to end at start of 2025).

```{r prev-census}

krt_census_prev <- raw_prev |> 
  mutate(date_end = case_when(is.na(date_death) ~ ymd(20250101),
                              TRUE ~ date_death)) |> 
  census_count(ymd(20180101), ymd(20250101),
                  "quarter",
                  date_start, date_end,
                  region)

krt_census_incid <- raw_incident |> 
  mutate(date_end = case_when(is.na(date_death) ~ ymd(20250101),
                              TRUE ~ date_death)) |> 
  census_count(ymd(20180101), ymd(20250101),
                  "quarter",
                  date_start, date_end,
                  region)

krt_census <- bind_rows(krt_census_prev, krt_census_incid) |> 
  group_by(date, region) |> 
  summarise(count = sum(prev)) |> 
  ungroup() 

krt_prev_census_eng <- krt_census |> 
  group_by(date) |> 
  summarise(count = sum(count)) |> 
  mutate(region = "England")

krt_census <- bind_rows(krt_census, krt_prev_census_eng)
 
  
```

## Prep for Stella

Take all flows and prevalence, convert to Stella var names and make wide.

```{r}

stella_prev_c <- krt_census |> 
  mutate(quart_start = quarter(date)/4 - 0.25,
         time = year(date) + quart_start) |> 
    filter(time >= 2020 & time < 2025) |> 
  mutate(stella_var = paste0("UKRR KRT[", region, "]")) |> 
  select(time, stella_var, count)

stella_incid_c <- incid_yr |>
  mutate(time = year(new_year)) |> 
    filter(time >= 2020) |> 
  mutate(stella_var = paste0("UKRR KRT incid[", region, "]")) |>
  select(time, stella_var, count = n)


  
stella_mort_c <- mort_yr |> 
  mutate(time = year(mort_year)) |> 
    filter(time >= 2020) |> 
  mutate(stella_var = paste0("UKRR KRT mort[", region, "]")) |> 
  select(time, stella_var, count = n)

stella_yr_ukrr_for_ckd <- bind_rows(stella_incid_c, stella_mort_c) |>
  pivot_wider(names_from = time, values_from = count) |>
  arrange(stella_var)

stella_prev_ukrr_for_ckd <- stella_prev_c |> 
  filter(time >= 2020) |> 
  pivot_wider(names_from = time, values_from = count) |>
  arrange(stella_var)


```

Calculate pay-off weights for use in Stella optimisation (based on Squared Error pay-off).

```{r}

prev_wt <- stella_prev_ukrr_for_ckd |> 
  rowwise() |> 
  mutate(wt = 1 / var(c_across("2020": "2024.75"), na.rm = TRUE))  |> 
  ungroup() |> 
  select(stella_var, wt)

inc_wt <- stella_yr_ukrr_for_ckd |> 
  rowwise() |> 
  mutate(wt = 1 / var(c_across("2020": "2024"), na.rm = TRUE))  |> 
  ungroup() |> 
  select(stella_var, wt)

wts <- bind_rows(prev_wt, inc_wt) |> 
  mutate(stella_var = str_replace(stella_var, fixed("["), " wt["))
```

Write to Excel.

```{r}

wb <- createWorkbook()  

addWorksheet(wb, "prev for ckd sd")
writeData(wb, "prev for ckd sd", stella_prev_ukrr_for_ckd)

addWorksheet(wb, "flows for ckd sd")
writeData(wb, "flows for ckd sd", stella_yr_ukrr_for_ckd)

addWorksheet(wb, "weights")
writeData(wb, "weights", wts)

saveWorkbook(wb, "ukrr census v2.xlsx", overwrite = TRUE)
```

Check stock and flow counts for validity.

```{r}

val_stock <- stella_prev_c |>
  mutate(region = str_extract(stella_var, "(?<=\\[).*?(?=\\])")) |> 
  rename(stock_start = count) |> 
    select(-stella_var)

val_flow_in <- incid_qrt_total |>
  mutate(time = year(new_quart) + quarter(new_quart)/4 - 0.25) |> 
  rename( flow_in = n) |> 
  select(- new_quart) |> 
  filter(time >= 2020)

val_flow_out <- mort_quart |> 
  mutate(time = year(mort_quarter) + quarter(mort_quarter)/4 - 0.25) |> 
  rename(flow_out = n) |>  
  select(- mort_quarter) |> 
  filter(time >= 2020)

val_check <- val_stock |> 
  full_join(val_flow_in, by = join_by(time, region)) |> 
  full_join(val_flow_out, by = join_by(time, region)) |> 
  arrange(region, time) |>
  mutate(stock_end = stock_start + flow_in - flow_out) |> 
  relocate(stock_start, .before = flow_in) |> 
  group_by(region) |> 
  mutate(net_flow = flow_in - flow_out,
         check = case_when(stock_end > lead(stock_start) ~ "high",
                           stock_end < lead(stock_start) ~ "low",
                           stock_end == lead(stock_start) ~ "match"),
         value = stock_end - lead(stock_start),
         pc = value / stock_end * 100) |> 
  ungroup()


wb <- createWorkbook()  

addWorksheet(wb, "stock flows")
writeData(wb, "stock flows", val_check)

saveWorkbook(wb, "ukrr valid check v2 2024.xlsx", overwrite = TRUE)
```
